---
title: "Guided Exercises for Chapter 5: Statistical Inference with Continuous Data (R Version) - Part 2"
author: ""
format: html
---

# Introduction

This set of guided exercises will help you work step-by-step through the statistical tests discussed in Sections 5.3 and 5.4 of Chapter 5. These exercises are designed to familiarize you with R and guide you through the analysis process using a practical, hands-on approach. Each exercise corresponds to the methods introduced in Sections 5.3 and 5.4 and provides detailed instructions on how to perform these analyses in R.

## Dataset Information

For these exercises, you will use the dataset `lowbwt.sav`, which you can download from Brightspace. The dataset is in SPSS format, so we will use the `haven` package to read it into R. This dataset contains information collected at Baystate Medical Center, Springfield MA, during 1986, and is from Appendix 1 of Hosmer and Lemeshow (1989). The data include information about factors related to low birth weight, which is an important concern due to its association with high infant mortality rates and birth defects.

The dataset includes the following variables:

- **ID**: Identification Code
- **lbw**: Low Birth Weight (0 = Birth Weight ≥ 2500g, 1 = Birth Weight < 2500g)
- **age**: Age of the Mother in Years
- **lwt**: Weight in Pounds at the Last Menstrual Period
- **ses**: Socio-Economic Status (1 = Low, 2 = Middle, 3 = High)
- **smoke**: Smoking Status During Pregnancy (1 = Yes, 0 = No)
- **ptl**: History of Premature Labor (0 = None, 1 = One, etc.)
- **ht**: History of httension (1 = Yes, 0 = No)
- **urirr**: Presence of Uterine Irritability (1 = Yes, 0 = No)
- **pvft**: Number of Physician Visits During the First Trimester (0 = None, 1 = One, 2 = Two, etc.)
- **bwt**: Birth Weight in Grams

## Exercise 1: Independent Samples t-Test in R

### Objective

Compare birth weights between mothers who smoked during pregnancy and those who did not to determine if there is a significant difference.

### Graphical Exploration

To get a sense of the data and the extent of a possible effect of smoking, we start by creating a boxplot. In R, this can be achieved by following these steps:

```{r}
# Load required packages
library(haven)
library(ggplot2)

# Load the dataset
lowbwt <- read_sav("lowbwt.sav")
lowbwt$ses <- lowbwt$race

# Create boxplot comparing birth weights for mothers who smoked and those who did not
ggplot(lowbwt, aes(x = factor(smoke, labels = c("No", "Yes")), y = bwt)) +
  geom_boxplot() +
  labs(x = "Smoking Status", y = "Birth Weight (grams)", title = "Birth Weight by Smoking Status")
```

::: {.callout-important icon=false title="Question"}
How do the spread, median, and potential outliers in the birth weight distributions compare between the two groups?
:::

### Analysis

```{r}
# Perform independent samples t-test
independent_t_test <- t.test(bwt ~ smoke, data = lowbwt, var.equal = TRUE)

# Print results
print(independent_t_test)
```

### Output Interpretation

The output from the `t.test()` function will include several key statistics:

- **t-value**: The value of the test statistic.
- **degrees of freedom (df)**: The degrees of freedom for the test.
- **p-value**: The probability of observing a test statistic as extreme as, or more extreme than, the observed value under the null hypothesis.
- **Confidence interval**: The range of values within which the true difference in means is likely to fall.

::: {.callout-important icon=false title="Question"}
Is there a significant difference in birth weights between mothers who smoked and those who did not?
:::

### Model Diagnostics

To assess whether the assumption of normality holds for the outcome variable in both groups, we can create histograms and perform a Shapiro-Wilk test:

```{r}
# Create histograms of birth weight by smoking status
lowbwt$smoke <- factor(lowbwt$smoke, labels = c("No", "Yes"))
ggplot(lowbwt, aes(x = bwt)) +
  geom_histogram(binwidth = 100, fill = "blue", alpha = 0.7) +
  facet_wrap(~smoke) +
  labs(x = "Birth Weight (grams)", title = "Histograms of Birth Weight by Smoking Status")

# Perform Shapiro-Wilk test for normality
shapiro_no <- shapiro.test(lowbwt$bwt[lowbwt$smoke == "No"])
shapiro_yes <- shapiro.test(lowbwt$bwt[lowbwt$smoke == "Yes"])

# Print results
print(shapiro_no)
print(shapiro_yes)
```

::: {.callout-important icon=false title="Question"}
Do the histograms and Shapiro-Wilk tests indicate that the birth weight data are approximately normally distributed for both groups?
:::

## Exercise 2: Mann-Whitney U Test in R

### Objective

Compare the distributions of birth weights between mothers with and without a history of httension using a non-parametric test.

### Graphical Exploration

To get a sense of the data and the extent of a possible effect of httension, we start by creating a boxplot. In R, this can be achieved by following these steps:

```{r}
# Create boxplot comparing birth weights for mothers with and without a history of httension
ggplot(lowbwt, aes(x = factor(ht, labels = c("No", "Yes")), y = bwt)) +
  geom_boxplot() +
  labs(x = "httension History", y = "Birth Weight (grams)", title = "Birth Weight by httension History")
```

### Analysis

```{r}
# Perform Mann-Whitney U Test
mann_whitney <- wilcox.test(bwt ~ ht, data = lowbwt)

# Print results
print(mann_whitney)
```

### Output Interpretation

The output from the `wilcox.test()` function will include:

- **W statistic**: The value of the test statistic.
- **p-value**: The probability of observing a test statistic as extreme as, or more extreme than, the observed value under the null hypothesis.

::: {.callout-important icon=false title="Question"}
Is there a significant difference in birth weights between mothers with and without a history of httension?
:::

## Exercise 3: One-Way ANOVA in R

### Objective

Compare birth weights across different socio-economic status groups to determine if there is a significant difference in means.

### Graphical Exploration

To get a sense of the data and the potential differences among socio-economic status groups, we start by creating a boxplot. In R, this can be achieved by following these steps:

```{r}
# Create boxplot comparing birth weights across socio-economic status groups
ggplot(lowbwt, aes(x = factor(ses, labels = c("Low", "Middle", "High")), y = bwt)) +
  geom_boxplot() +
  labs(x = "Socio-Economic Status", y = "Birth Weight (grams)", title = "Birth Weight by Socio-Economic Status")
```

### Analysis

```{r}
# Perform One-Way ANOVA
anova_result <- aov(bwt ~ factor(ses), data = lowbwt)

# Print summary of ANOVA
summary(anova_result)
```

### Output Interpretation

The output from the `aov()` function will include:

- **F-statistic**: The value of the test statistic.
- **p-value**: The probability of observing a test statistic as extreme as, or more extreme than, the observed value under the null hypothesis.

::: {.callout-important icon=false title="Question"}
Is there a significant difference in birth weights across the socio-economic status groups?
:::

## Exercise 4: Kruskal-Wallis Test in R

### Objective

Compare the distributions of birth weights across different socio-economic status groups using a non-parametric test.

### Graphical Exploration

To get a sense of the data and the potential differences among socio-economic status groups, we start by creating a boxplot. In R, this can be achieved by following these steps:

```{r}
# Create boxplot comparing birth weights across socio-economic status groups
ggplot(lowbwt, aes(x = factor(ses, labels = c("Low", "Middle", "High")), y = bwt)) +
  geom_boxplot() +
  labs(x = "Socio-Economic Status", y = "Birth Weight (grams)", title = "Birth Weight by Socio-Economic Status")
```

### Analysis

```{r}
# Perform Kruskal-Wallis Test
kruskal_result <- kruskal.test(bwt ~ factor(ses), data = lowbwt)

# Print results
print(kruskal_result)
```

### Output Interpretation

The output from the `kruskal.test()` function will include:

- **Kruskal-Wallis chi-squared**: The value of the test statistic.
- **p-value**: The probability of observing a test statistic as extreme as, or more extreme than, the observed value under the null hypothesis.

::: {.callout-important icon=false title="Question"}
Are there significant differences in birth weights between the socio-economic status groups?
:::