---
title: "Medical Statistics - Answers lab 5"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
  pdf:
    toc: false
execute:
  warning: false
  message: false
  error: false
  eval: true
---

## Introduction

This answer page gives one complete set of answers for lab 5. The same numerical results should be obtained when the analyses are carried out in R or SPSS.

The analyses use the Cleveland heart disease dataset and address the following research questions:

* Does maximum heart rate achieved differ across chest pain types?
* Is the presence of heart disease associated with sex?
* Is the presence of heart disease associated with fasting blood sugar?

## Exercise 1: Download and open the dataset

The dataset contains 303 patients and 14 variables.

```{r}
library(haven)
library(knitr)
library(tableone)
library(car)
library(dunn.test)

heart_data <- as.data.frame(read_sav("datasets/heart_disease_cleveland.sav"))

# Convert labelled categorical variables to factors.
labelled_vars <- vapply(heart_data, haven::is.labelled, logical(1))
heart_data[labelled_vars] <- lapply(heart_data[labelled_vars], haven::as_factor)

dim(heart_data)
names(heart_data)
```

The variables `sex`, `cp`, `fbs`, `restecg`, `exang`, `slope`, `thal`, and `target` are categorical. The variable `ca` is also best treated as categorical for the baseline table, because it records the number of major vessels from 0 to 3.

```{r}
heart_data$ca <- factor(heart_data$ca)
```

## Exercise 2: Create a baseline characteristics table

```{r}
continuous_vars <- c("age", "trestbps", "chol", "thalach", "oldpeak")
categorical_vars <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "ca", "thal")
all_vars <- c(continuous_vars, categorical_vars)

table1 <- CreateTableOne(
  vars = all_vars,
  data = heart_data,
  factorVars = categorical_vars
)

kable(
  print(
    table1,
    nonnormal = "oldpeak",
    printToggle = FALSE
  )
)
```

A suitable baseline table gives one summary format per variable. In the table above, most continuous variables are summarized using mean and standard deviation, `oldpeak` is summarized using median and interquartile range because it is clearly skewed, and categorical variables are summarized using counts and percentages.

## Exercise 3: Analyze maximum heart rate across chest pain types

Research question: *Does maximum heart rate achieved differ across chest pain types?*

The outcome is `thalach`, and the grouping variable is `cp`.

### Descriptive results

```{r}
boxplot(
  thalach ~ cp,
  data = heart_data,
  xlab = "Chest pain type",
  ylab = "Maximum heart rate achieved",
  las = 2
)
```

```{r}
for (group in levels(heart_data$cp)) {
  group_data <- heart_data$thalach[heart_data$cp == group]
  n <- length(group_data)
  mean_thalach <- mean(group_data)
  se_thalach <- sd(group_data) / sqrt(n)
  ci_thalach <- mean_thalach + c(-1, 1) * qt(0.975, df = n - 1) * se_thalach

  cat(
    group,
    ": n =", n,
    ", mean =", round(mean_thalach, 1),
    ", 95% CI = [", round(ci_thalach[1], 1), ", ", round(ci_thalach[2], 1), "]\n",
    sep = ""
  )
}
```

The mean maximum heart rate was lowest in the asymptomatic chest pain group.

### One-way ANOVA

```{r}
anova_model <- aov(thalach ~ cp, data = heart_data)
summary(anova_model)
```

The one-way ANOVA is statistically significant, $F(3, 299) = 17.41$, $p < 0.001$. This indicates that maximum heart rate achieved differs across chest pain types.

### Assumption checks

```{r}

par(mfrow = c(2, 2))
hist(heart_data$thalach[heart_data$cp == "Typical angina"],
     main = "Typical angina", xlab = "Maximum heart rate")
hist(heart_data$thalach[heart_data$cp == "Atypical angina"],
     main = "Atypical angina", xlab = "Maximum heart rate")
hist(heart_data$thalach[heart_data$cp == "Non-anginal pain"],
     main = "Non-anginal pain", xlab = "Maximum heart rate")
hist(heart_data$thalach[heart_data$cp == "Asymptomatic"],
     main = "Asymptomatic", xlab = "Maximum heart rate")

par(mfrow = c(1, 1))
qqnorm(residuals(anova_model))
qqline(residuals(anova_model))

tapply(heart_data$thalach, heart_data$cp, sd)
leveneTest(thalach ~ cp, data = heart_data)
```

The histograms and Q-Q plot can be used to assess whether the distributions are roughly compatible with ANOVA. There is no need to base this decision on a formal test of normality; with these sample sizes such tests are often not helpful for deciding which analysis to use. The group standard deviations are of a similar order of magnitude, and Levene's test does not show evidence against the equal variances assumption.

### Post-hoc comparisons

```{r}
pairwise.t.test(
  heart_data$thalach,
  heart_data$cp,
  p.adjust.method = "bonferroni",
  pool.sd = TRUE
)
```

Using Bonferroni-adjusted pairwise t-tests, the asymptomatic chest pain group has a significantly lower maximum heart rate than the typical angina group, the atypical angina group, and the non-anginal pain group. The other pairwise comparisons are not statistically significant after Bonferroni adjustment.

### Kruskal-Wallis sensitivity analysis

```{r}
kruskal.test(thalach ~ cp, data = heart_data)

dunn.test(
  heart_data$thalach,
  heart_data$cp,
  method = "bonferroni",
  altp = TRUE
)
```

The Kruskal-Wallis test gives the same substantive conclusion, $\chi^2(3) = 47.32$, $p < 0.001$. The Bonferroni-adjusted Dunn's test again shows that the asymptomatic group differs from the other three chest pain groups.

## Exercise 4: Analyze heart disease by sex

Research question: *Is the presence of heart disease associated with sex?*

::: {.callout-note}
For the 2x2 tables in this lab, `chisq.test(..., correct = FALSE)` is used to report the ordinary Pearson chi-square test. This corresponds to the **Pearson Chi-Square** row in SPSS. If `correct = FALSE` is omitted, R applies Yates' continuity correction by default for 2x2 tables; SPSS reports this separately as **Continuity Correction**.
:::

```{r}
sex_table <- table(heart_data$sex, heart_data$target)
sex_table
round(prop.table(sex_table, margin = 1) * 100, 1)
```

The row percentages show that the prevalence of heart disease was higher in males than in females.

The expected cell counts are all larger than 5, so the chi-square test is appropriate.

```{r}
sex_chisq <- chisq.test(sex_table, correct = FALSE)
sex_chisq$expected
sex_chisq
```

There is strong evidence of an association between sex and heart disease, $\chi^2(1) = 23.22$, $p < 0.001$. The prevalence of heart disease was higher in males (55.3%) than in females (25.8%).

## Exercise 5: Analyze heart disease by fasting blood sugar

Research question: *Is the presence of heart disease associated with fasting blood sugar?*

```{r}
fbs_table <- table(heart_data$fbs, heart_data$target)
fbs_table
round(prop.table(fbs_table, margin = 1) * 100, 1)
```

The row percentages show similar heart disease prevalence in the two fasting blood sugar groups.

The expected cell counts are all larger than 5, so the chi-square test is appropriate.

```{r}
fbs_chisq <- chisq.test(fbs_table, correct = FALSE)
fbs_chisq$expected
fbs_chisq
```

There is no evidence of an association between fasting blood sugar and heart disease, $\chi^2(1) = 0.19$, $p = 0.660$. The prevalence of heart disease was similar for patients with fasting blood sugar at or below 120 mg/dl (45.3%) and patients with fasting blood sugar above 120 mg/dl (48.9%).
